---
title: "Typography & Annotations"
---

# 6.1 Typography

## Choosing Typeface & Fonts

-   context matters
-   when in doubt, use a sans-serif style
    -   styles with no edges\\
-   when adding numbers, choose a mono spaces typeface with figures that ...

## Fonts in `ggplot`

-   packages

    -   `showtext`

    -   `extrafont`

Example - update fonts on dumbbell plot

```{r, include=FALSE}
# wrangle data

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(showtext)

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                wrangle data                              ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

jobs_clean <- jobs |>

  # add cols (needed for dumbbell plot) ----
  mutate(percent_male = 100 - percent_female, # % of females within each industry was already included
       difference_earnings = total_earnings_male - total_earnings_female) |>  # diff in earnings between M & F

  # rearrange columns ----
  relocate(year, major_category, minor_category, occupation,
           total_workers, workers_male, workers_female,
           percent_male, percent_female,
           total_earnings, total_earnings_male, total_earnings_female, difference_earnings,
           wage_percent_of_male) |>

  # drop rows with missing earning data ----
  drop_na(total_earnings_male, total_earnings_female) |>

  # make occupation a factor ----
  mutate(occupation = as.factor(occupation)) |>

# ---- this next step is for creating our dumbbell plots ----

# classify jobs by percentage male or female ----
  mutate(group_label = case_when(
    percent_female >= 75 ~ "Occupations that are 75%+ female",
    percent_female >= 45 & percent_female <= 55 ~ "Occupations that are 45-55% female",
    percent_male >= 75 ~ "Occupations that are 75%+ male"
  ))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                              create subset df                            ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#....guarantee the same random samples each time we run code.....
set.seed(0)

#.........get 10 random jobs that are 75%+ female (2016).........
f75 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 75%+ female") |>
  slice_sample(n = 10)

#..........get 10 random jobs that are 75%+ male (2016)..........
m75 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 75%+ male") |>
  slice_sample(n = 10)

#........get 10 random jobs that are 45-55%+ female (2016).......
f50 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 45-55% female") |>
  slice_sample(n = 10)

#.......combine dfs & relevel factors (for plotting order).......
subset_jobs <- rbind(f75, m75, f50) |>
  mutate(group_label = fct_relevel(group_label, 
                                   "Occupations that are 75%+ female",
                                   "Occupations that are 45-55% female", 
                                   "Occupations that are 75%+ male"))
```

Prepare Customizations

```{r}
# create palette w/colors for male vs female, and text
earnings_pal <- c("males" = "#2D7787",
                  "females" = "#FC6B4B",
                  "dark_text" = "#0C1509",
                  "light_text" = "#4E514D")

# can use view_palette functin to see preview of pal
monochromeR::view_palette(earnings_pal)
```

```{r}
# import Google fonts
font_add_google(name = "Josefin Sans", # name as it is on Google
                family = "josefin") # custom name we'll use to call the font

font_add_google(name = "Sen", 
                family = "sen")

# enable showtext ----
showtext_auto() #helps keep fonts consistent across devices

# import Font Awesome fonts
font_add(family = "fa-brands",
         regular = here::here("fonts", "Font Awesome 6 Brands-Regular-400.otf"))
font_add(family = "fa-regular",
         regular = here::here("fonts", "Font Awesome 6 Free-Regular-400.otf")) 
font_add(family = "fa-solid",
         regular = here::here("fonts", "Font Awesome 6 Free-Solid-900.otf"))
```

Apply new color to dumbell plot

```{r}
# store plot to add themes separately
plot <- ggplot(subset_jobs) +
geom_segment(aes(x = total_earnings_female, xend = total_earnings_male,
                 y = fct_reorder(occupation, total_earnings), yend = occupation)) +
  geom_point(aes(x = total_earnings_male, y = occupation),
             color = earnings_pal["males"], size = 3.25) +
  geom_point(aes(x = total_earnings_female, y = occupation),
             color = earnings_pal["females"], size = 3.25) +
  facet_wrap(~group_label, nrow = 3, scales = "free_y") +
  scale_x_continuous(labels = scales::label_dollar(scale = 0.001, suffix = "k"),
                     breaks = c(25000, 50000, 75000, 100000, 125000))
```

```{r}
plot +
  labs(title = "Males earn more than gemales across most occupations",
       subtitle = "Median earnings of full-time <span style = 'color:#2D7787; font-size: 20pt;'>**male**</span> vs. <span style = 'color:#FC6B4B; font-size: 20pt;'>**female**</span> workers by occupation in 2016",
       caption = "Data Source: TidyTuesday (March 5, 2019)") +
  
  # apply theme minimal to simplify theme & then build off it
  theme_minimal() +
  
  # Update theme (text and axes)
  theme(
    
    # adjust title and subtitle to be on the left
    plot.title.position = "plot",
    
    # update title size, face, and to a color in our pal
    plot.title = element_text(face = "bold",
                              family = "josefin",
                              size = 25,
                              color = earnings_pal["dark_text"]),
    
    # update subtitle size, color, and spacing
    plot.subtitle = ggtext::element_textbox_simple(size = 17,
                                 family = "sen",
                                 color = earnings_pal["light_text"],
                                 margin = margin(0.5, 0, 1, 0,
                                                 unit = "lines")),
    # update caption size, color, and spacing
    plot.caption = element_text(face = "italic",
                                family = "sen",
                                color = earnings_pal["light_text"],
                                margin = margin(4, 0, 0, 0, unit = "lines")),
    
    # update text above each panel (facet)
    strip.text.x = element_text(face = "bold",
                                family = "josefin",
                                size = 12,
                                hjust =0),
    
    # update panel spacing (more space b/w panels)
    panel.spacing.y = unit(1, "lines"),
    
    # update ALL axis text (occupations listed and $ values)
    axis.text = element_text(color = earnings_pal["light_text"],
                             family = "sen"),
    
    # make x-axis text larger than y-axis
    axis.text.x = element_text(size = 10),
    
    # remove axis titles
    axis.title = element_blank()
  )
```

# 6.2 Annotations

## Examples

### Temperature on Lobster Metabolism

Plot code to start

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#.........................load libraries.........................
library(tidyverse)

#..........................read in data..........................

# read in Google Sheet ----
lobs <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1DkDVcl_9rlaqznHfa_v1V1jtZqcuL75Q6wvAHpnCHuk/edit#gid=2143433533") |>
  mutate(temp = as.factor(temp))

# # alternatively, read in csv file ----
# lobs <- read_csv(here::here("week6", "data", "metabolism-foraging-data.csv")) |>
#   mutate(temp = as.factor(temp))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            create lobster plot                           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................create theme..........................
lob_theme <- function(){
  theme_light() +
    theme(
      axis.title.x = ggtext::element_markdown(size = 13,
                                              margin = margin(t = 1, r = 0, b = 0, l = 0, unit = "lines")),
      axis.title.y = ggtext::element_markdown(size = 13,
                                              margin = margin(t = 0, r = 1, b = 0, l = 0, unit = "lines")),
      axis.text = element_text(color = "black", size = 12),
      panel.border = element_rect(colour = "black", size = 0.7),
      panel.grid = element_blank(),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      legend.position = c(0.95, 0.95),
      legend.justification = c(0.95, 0.95),
      legend.box.background = element_rect(color = "black", size = 1.1)

    )
}

#..........................create scales.........................
lob_palette <- c("11" = "#7B8698",
                 "16" = "#BAD7E5",
                 "21" = "#DC7E7C",
                 "26" = "#7D3E40")

lob_shapes <-  c("11" = 15,
                 "16" = 16,
                 "21" = 17,
                 "26" = 18)

lob_sizes <- c("11" = 6,
               "16" = 6,
               "21" = 6,
               "26" = 7) #diamond looks smaller than other shapes, increase size

#........................create plot text........................
x_axis_lab <- glue::glue("Resting Metabolic Rate<br>
                         (mg O<sub>2</sub> kg<sup>-1</sup> min<sup>-1</sup>)")

y_axis_lab <- glue::glue("Maximum Consumption Rate<br>
                         (prey consumed predator<sup>-1</sup> 24hr<sup>-1</sup>)")

#............................plot data...........................
lob_plot <- ggplot(lobs, aes(x = SMR, y = avg_eaten,
                 color = temp, shape = temp, size = temp)) +
  geom_point() +
  scale_color_manual(values = lob_palette, name = "Temperature (ºC)") +
  scale_shape_manual(values = lob_shapes, name = "Temperature (ºC)") +
  scale_size_manual(values = lob_sizes, name = "Temperature (ºC)") +
  scale_x_continuous(breaks = seq(0, 1.5, by = 0.2)) +
  scale_y_continuous(breaks = seq(0, 35, by = 5)) +
  labs(x = x_axis_lab,
       y = y_axis_lab) +
  lob_theme()

```

Building custom annotations

-   `geom_text()` and `geom_label()` take aesthetic mappings, draw the geom once per each row of the data frame

```{r}
lob_plot +
  
  # add text for important lobsters
  geom_text(
    x = 0.1, y = 24, # use coords to place text
    label = "Important Lobsters",
    size = 4, 
    color = "black",
    hjust = "inward"
  ) +
  
  # add bounding box for important lobsters
  geom_rect(
    xmin = 0.25, xmax = 0.85, # min and max values for box positions
    ymin = 8.5, ymax = 18,
    alpha = 0.5, # translucent
    fill = "gray40", # fill rect 
    color = "black", # box border
    show.legend = FALSE # remove border around points in legend
  )
```

-   `anotate()`
    -   better option
```{r}
lob_plot +
  
  # add text for important lobsters
  geom_text(
    x = 0.1, y = 24, # use coords to place text
    label = "Important Lobsters",
    size = 4, 
    color = "black",
    hjust = "inward"
  ) +
  
  # add bounding box for important lobsters
  annotate(
    geom = "rect",
    xmin = 0.25, xmax = 0.85, # min and max values for box positions
    ymin = 8.5, ymax = 18,
    alpha = 0.5, # translucent
    fill = "gray40", # fill rect 
    color = "black" # box border
  ) +
  
  # add arrow from text and box
  annotate(
    geom = "curve",
    x = 0.3, xend = 0.5, # define where the arrow starts and ends
    y = 23.8, yend = 19,
    curvature = -0.15, # change direction of curve
    arrow = arrow(length = unit(0.3, "cm")) # add points on arrow
  )
```

