---
title: 'Week 3 Discussion'
editor_options: 
  markdown: 
    wrap: 72
---

# Key Plot Types

## Histogram vs. Density

-   Density
    -   area under curve should equal 1
    -   shows the probability of getting a particular x-value
-   Histogram
    -   area under curve does NOT equal 1
    -   y-axis is the count, how many times a particular x-value appears

## Ridgeline plots

-   when they're most useful
    -   for comparing distributions of different groups
    -   looking at different distributions over time (timeseries)
-   unlike density plot, ridgeline does not have a numeric y-axis

## Boxplot vs Violin plot

-   Boxplot
    -   shows key summary statistics

        -   median, 25-50-75 quantiles
-   violin plot
    -   shows distribution, where wider points are areas with more
        common values
-   overlaying the two allows you to view distribution AND summary
    statistics at the same time
    -   recommended/common

# Drought Exercise

### Preparation

```{r}
# load libraries
library(tidyverse)
library(ggridges)

# import data
tuesdata <- tidytuesdayR::tt_load('2021-07-20')
drought <- tuesdata$drought
```

```{r}
# create an ugly plot
ggplot(drought, aes(x = area_total, y = state_abb)) +
  geom_density_ridges2()
```

### Wrangling

Consider why each step is important

```{r}
# wrangle data
drought_clean <- drought %>% 

  # select cols of interest & update names as needed ----
  select(date = valid_start, state_abb, drought_lvl, area_pct) %>% 

  # add year, month & day cols using {lubridate} fxns ----
  mutate(year = year(date),
         month = month(date, label = TRUE, abbr = TRUE),
         day = day(date)) %>% 

  # add drought level conditions names ----
  mutate(drought_lvl_long = factor(drought_lvl,
                            levels = c("D4", "D3", "D2", "D1","D0", "None"),
                            labels = c("(D4) Exceptional", "(D3) Extreme",
                                       "(D2) Severe", "(D1) Moderate", "(D0) Abnormally Dry", 
                                       "No Drought"))) %>% 
  
  # reorder cols ----
  select(date, year, month, day, state_abb, drought_lvl, drought_lvl_long, area_pct)

```

### Plots

Stacked area plot of drought conditions through time

```{r}
drought_clean %>% 
  
  # remove drought_lvl "None" & filter for just CA ----
  filter(drought_lvl != "None",
         state_abb == "CA") %>% 
  
  # create ggplot ----
  ggplot(mapping = aes(x = date, y = area_pct, fill = drought_lvl_long)) +
  
  # reverse order of groups so level D4 is closest to x-axis ----
  geom_area(position = position_stack(reverse = TRUE)) +
  
  # update colors to match US Drought Monitor (colors selected using ColorPick Eyedropper from original USDM data viz) ----
  scale_fill_manual(values = c("#853904", "#FF0000", "#FFC100", "#FFD965", "#FFFF00")) +
  
  # set x-axis breaks & remove padding between data and x-axis ----
  scale_x_date(breaks = scales::breaks_pretty(n = 10),
               expand = c(0, 0)) +

  # set y-axis breaks & convert values to percentages & & remove padding between data and y-axis----
  scale_y_continuous(breaks = seq(0, 100, by = 10),
                     labels = scales::label_percent(scale = 1),
                     expand = c(0, 0)) +
  
  # add title ----
  labs(title = "Drought area in California") 
```
