---
title: "Colors and Choropleths"
subtitle: "Combined lecture notes, 5.1-5.3"
author: "Briana Barajas"
date: 2024-02-05
format:
  html:
    code-fold: true
    code-summary: "View code"
toc: TRUE
---

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, include=FALSE, results='hide'}
library(palmerpenguins)
library(tidyverse)
library(paletteer)
library(tigris)
```

## 5.2 Colors

### Key Considerations

-   key questions while making visualizations
-   is using color the best and/or only way to visually represent your variable(s)?
-   are you designing for a particular organization/brand?
-   what emotions are you trying to elicit or avoid?
-   who's the intended audience?
-   picking colors (summary)
-   ensure you're picking color-blind friendly palettes
-   use color wheels to identify color harmonies
-   think carefully about what emotions/messages your color choices will convey
-   avoid lots of pure, fully-saturated hues

### Basics about color

-   categorical

-   mainly formed by selecting differet *hues*

-   hues assigned to each group must be distinct and ideally have different lightnesses

-   groups don't have an intrinsic order

-   limit to no more than 7 hues

-   sequential scales

-   colors assigned to data values in a continuum, based on lightness, hue, or both

-   lower values typically associated with lighter colors & higher values associated with darker colors

-   can use a single hue *or* two hues

-   diverging scales

-   combination of two sequential palettes with a shared endpoint at the central value

-   central value is assigned a light color (light gray is best)

-   use a distinctive hue for each of the component palettes

-   Example

```{r, warning=FALSE}
# store categorical color plot
cat_color_plot <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species, shape = species)) +
  geom_point(size = 4, alpha = 0.8)

cat_color_plot 
```

```{r, warning=FALSE}
# store continuous color scale plot
cont_color_plot <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = body_mass_g)) +
  geom_point(size = 4, alpha = 0.8) 

cont_color_plot 
```

### Color blind friendly

-   Rainbow color maps

-   short falls

-   horrible for printing (gray scale)

-   not color-blind friendly (looks like two colors)

-   no natural order, no color is inherently higher or lower

-   alternatives, some attempts to remedy the problem

-   view slide 13/63 for improved colormaps

-   `viridis` package

-   offers continuous and binned color scales

-   can tell difference in lightness even in gray scale

-   more uniform than standard rainbow color scales

-   Example: applying `viridis` color schemes

```{r, warning=FALSE}
# add standard viridis color scale to continuous variable
cat_color_plot +
  scale_color_viridis_d(option = "viridis")

# select continous scale from viridis package
cont_color_plot +
  scale_color_viridis_c(option = "magma")
```

-   can also use `RColorBrewer`

-   not all color by friendly, but they can be filtered

-   Example: applying `RColorBrewer` color schemes

```{r, warning=FALSE}
# apply categorical color scheme
cat_color_plot +
  scale_color_brewer(palette = "Dark2") 

# apply continuous color scheme
cont_color_plot +
  scale_color_distiller(palette = "BuPu")

# classed versions of continuous color scales
cont_color_plot +
  scale_color_fermenter(palette = "YlGnBu")
```

-   can highlight points to make colors more visible

```{r, warning=FALSE}
# plot bill length vs. depth
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                     fill = body_mass_g)) +
  geom_point(shape = 21, size = 4, alpha = 0.8) +
  scale_fill_distiller(palette = "BuPu")
```

### `{paletteer}` package

-   interface for accessing new-comprehensive list of color palettes

-   can generate a list of HEX codes

-   can apply a palette directly to `ggplot`

-   using `scale_*paletteer*()`

```{r, warning=FALSE}
# use paletteer for categorical plot
cat_color_plot +
  paletteer::scale_color_paletteer_d("calecopal::superbloom3")
```

### Additional Color Practice

-   Apply a custom palette to `ggplot`

```{r, warning=FALSE}
# create custom palette
my_palette <- c("#32DE8A", "#E36414", "#0F4C5C")

# apply to categorical plot
cat_color_plot +
  scale_color_manual(values = my_palette)
```

-   assigning colors to names/categories

```{r, warning=FALSE}
# create a named color palette (associate spp and color)
my_palette_named <- c("Adelie" = "#32DE8A","Chinstrap" = "#E36414", "Gentoo" = "#0F4C5C")

# now if a group is removed, the spp color remains the same
penguins %>% 
  filter(species != "Chinstrap") %>%  
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_palette_named)
```

-   modify data frame to apply colors to observations

```{r, warning=FALSE}
penguins %>% 
  mutate(
    my_color = case_when(
      bill_length_mm <40 ~ "#D7263D",
      between(bill_length_mm, 40,50) ~ "#E4BB97",
      bill_length_mm > 50 ~ "#386150") # END case-when
  ) %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = my_color)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_identity()
```

-   modify data frame to highlight select observations

```{r, warning=FALSE}
penguins %>% 
  mutate(
    my_color = case_when(
      body_mass_g > 6000 ~ "#D7263D",
      TRUE ~ "gray50"
    )
  ) %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = my_color)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_identity(guide = "legend", 
                       name = "Body mass (g)", 
                       labels = c(">6000", "<= 6000"))

```

# 5.3 Choropleths

### Introduction

choropleths are maps that **display the spatial distribution of a variable** across divided geohgraphical regions, where the **variable is encoded by color**

### Example - Cloropleth Map

Data wrangling

```{r, results='hide'}

#.........................get shape data.........................
county_geo <- tigris::counties(class = "sf", cb = TRUE) |> # cb = TRUE to use cartographic boundary files
  
  # shift US to fit AK, HI, PR (we'll be filtering these out though) and transform CRS to USA Contiguous Albers Equal Area Conic (ESRI:102003) ----
  shift_geometry()

#....................import precipitation data...................
precip_data <- read_csv(here::here("week5", "data", "county-jan19-dec23-precip.csv"), skip = 4)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ wrangle geometries  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~

county_geo_wrangled <- county_geo |>
  
  # clean up col names ----
  janitor::clean_names() |>
  
  # rename county & state cols ----
  rename(county = namelsad, state = state_name) |>
  
  # remove states / territories that we don't have precip data for ----
  filter(!state %in% c("Alaska", "Hawaii", "District of Columbia",
                       "United States Virgin Islands", "Puerto Rico", "American Samoa",
                       "Commonwealth of the Northern Mariana Islands", "Guam")) |>
  
  # capitalize "city" (VA) ----
  mutate(county = str_replace(string = county, pattern = " city", replacement = " City"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ wrangle precipitation data  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

precip_wrangled <- precip_data |>
  
  # clean up col names ----
  janitor::clean_names() |>
  
  # rename county col ----
  rename(county = name) |>
  
  # filter out DC ----
  filter(!county %in% c("Washington, D.C.")) |>
  
  # update name to match that in county_geo df ----
  mutate(county = str_replace(string = county, pattern = "Dona Ana County", replacement = "Doña Ana County")) |>
  
  # coerce precip & 20th centruy avg from chr to numeric ----
  mutate(value = as.numeric(value),
         x1901_2000_mean = as.numeric(x1901_2000_mean)) |>
  
  # calculate % change ----
  mutate(perc_change = ((value - x1901_2000_mean)/x1901_2000_mean)*100) |>
  
  # select, rename, reorder cols ----
  select(id, state, county, mean_1901_2000 = x1901_2000_mean, precip = value, perc_change, anomaly_1901_2000_base_period)

##~~~~~~~~~~~~~~~~~~
##  ~ join dfs  ----
##~~~~~~~~~~~~~~~~~~

# join dfs (be sure to join precip TO sf object, not the other way around) -------
joined_precip_geom <- full_join(county_geo_wrangled, precip_wrangled) 
```

Creating a base map

```{r}
# store map
base_map <- ggplot(joined_precip_geom) +
  
  # set aesthetics in geom layer (could do it in ggplot)
  geom_sf(aes(fill = perc_change),
          
          # minimize graticules
          linewidth = 0.1) +
  
  # add labels
  labs(title = "5-year precipitation compared to 20th century average",
       subtitle = "January 2019 - December 2023",
       caption = "Source: National Centers for Environmental Information") +
  
  # change theme to minimize non-data ink
  theme_void() +
  
  # continue customizations, 
  theme(legend.position = "bottom", # change legend position
        legend.title = element_blank(), # remove legend title
        plot.caption = element_text(face = "italic",
                                    margin = margin(t = 2, r = 0.5,
                                                    b = 0, l = 0,
                                                    "lines")))

# plot map
base_map
```

Continuous chloropleth map

```{r}
# store color paletteto use on map (n = number of colors)
my_brew_palette11 <- RColorBrewer::brewer.pal(n = 11, name = 'BrBG')

# update map
base_map +
  # apply continuous color gradient
  scale_fill_gradientn(colors = my_brew_palette11,
                       # adjust labels
                       labels = scales::label_percent(scale = 1),
                       breaks = scales::breaks_width(10),
                       values = scales::rescale(x = c(
                         min(na.omit(joined_precip_geom)$perc_change),
                         max(na.omit(joined_precip_geom)$perc_change))
                       )) +
   
  # update legend, make bar longer
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75))
```

Classed map

```{r}
# update color pal
my_brew_palette10 <- RColorBrewer::brewer.pal(n = 10, name = 'BrBG')

# update basemaps
base_map +
  
  # applybard width and wreight 
  scale_fill_stepsn(colors = my_brew_palette10,
                    values = scales::rescale(x = c(
                         min(na.omit(joined_precip_geom)$perc_change),
                         max(na.omit(joined_precip_geom)$perc_change))
                       )) +
    
    guides (fill = guide_colorsteps(barwidth = 25, barheight = 0.75),
            )

```
